---
title: "Final Project"
author: "Oghosa"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(haven) 
library(readr) 
library(dplyr) 
library(tableone)
<<<<<<< HEAD
library(ResourceSelection)
library(lmtest)
library(car)
library(ggplot2)
library(broom)

pacman::p_load(haven, readr, dplyr, tidyverse, table1, DiagrammeR, rsvg, DiagrammeRsvg, readxl, tableone, ResourceSelection, lmtest, car)
=======
pacman::p_load(haven, readr, dplyr, tidyverse, table1, DiagrammeR, rsvg, DiagrammeRsvg, readxl, tableone)
>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
```

```{r}
#Study Objectives: a) Assess the relationship between physical activity and the likelihood of reporting excellent general health among individuals aged 30 and above. b) Evaluate whether BMI and age act as confounding factors in the relationship between physical activity and excellent general health.

# This study will utilize data from the 2023 Behavioral Risk Factor Surveillance System (BRFSS). 

#Load dataset
fp <- read_xpt("D:\\ADA\\Final Project\\LLCP2023XPT\\LLCP2023.XPT")
```

```{r}
#Check data in age variable
table(fp$`_AGEG5YR`, useNA = "ifany")
```

```{r}
#From the data dictionary, age category '14' represents "Don't know/Refused/Missing". We remove this since it does not contribute to our analysis
fp1 <- fp %>%
  filter(`_AGEG5YR` != 14)
```

```{r}
#Recheck data in age variable to ensure '14' has been filtered out
table(fp1$`_AGEG5YR`, useNA = "ifany")
```

```{r}
#Recode the various age groups to match labels in the data dictionary
fp2 <- fp1 %>%
  mutate(Age_Group = case_when(
    `_AGEG5YR` == 1 ~ "18-24",
    `_AGEG5YR` == 2 ~ "25-29",
    `_AGEG5YR` == 3 ~ "30-34",
    `_AGEG5YR` == 4 ~ "35-39",
    `_AGEG5YR` == 5 ~ "40-44",
    `_AGEG5YR` == 6 ~ "45-49",
    `_AGEG5YR` == 7 ~ "50-54",
    `_AGEG5YR` == 8 ~ "55-59",
    `_AGEG5YR` == 9 ~ "60-64",
    `_AGEG5YR` == 10 ~ "65-69",
    `_AGEG5YR` == 11 ~ "70-74",
    `_AGEG5YR` == 12 ~ "75-79",
    `_AGEG5YR` == 13 ~ "80 or older"
  ))
```

```{r}
#Check to ensure transformation was successful
table(fp2$Age_Group, useNA = "ifany")
```

```{r}
#Filter out age categories that are less than 30 since my research question is focused on individuals that are aged 30 and above
fp3 <- fp2 %>%
  filter(Age_Group >= 30)
```

```{r}
# Recode BMI categories

fp3 <- fp3 %>%
  mutate(BMI_category = case_when(
    `_BMI5CAT` == 1 ~ "Underweight",
    `_BMI5CAT` == 2 ~ "Normal",
    `_BMI5CAT` == 3 ~ "Overweight",
    `_BMI5CAT` == 4 ~ "Obese"
  ))

# Recode BMI categories. 
#Ignore entries that do not contribute to our analysis (refused, not asked/missing, don't know or not sure)

fp3 <- fp3 %>%
  mutate(Physical_Activity = case_when(
    EXEROFT1 >= 101 & EXEROFT1 <= 199 ~ "Times per week",
    EXEROFT1 >= 201 & EXEROFT1 <= 299 ~ "Times per month",
  ))
```

```{r}
#Check if properly recoded
table(fp3$BMI_category, useNA = "ifany")
table(fp3$Physical_Activity, useNA = "ifany")
```

```{r}
# Remove rows with missing values in BMI_category and Physical_Activity
fp4 <- fp3 %>%
  filter(!is.na(BMI_category), !is.na(Physical_Activity))

```

```{r}
#Check if removal of missing values was successful
table(fp4$BMI_category, useNA = "ifany")
table(fp4$Physical_Activity, useNA = "ifany")
```

```{r}
#Re-coding General health status and creating a binary variable of health status while removing missing values and instances like refused, not sure, don't know, etc. per the dictionary
fp5 <- fp4 %>%
  mutate(GENHLTH_binary = case_when(
    GENHLTH == 1 ~ 1,   # Excellent health status
    GENHLTH %in% c(2, 3, 4, 5) ~ 0   # Very good, Good, Fair, Poor
  )) %>%
  filter(!is.na(GENHLTH), !GENHLTH %in% c(7, 9))  # Remove rows with 7, 9, or NA
```

```{r}
#Checking the tables for correct output
table(fp5$Physical_Activity, useNA = "ifany")
table(fp5$GENHLTH_binary, useNA = "ifany")
```

```{r}
# Summarize each variable individually
age_group_table <- fp5 %>%
  count(Age_Group) %>%
  mutate(
    Variable = "Age_Group",
    Proportion = paste0(round(100 * n / sum(n), 1), "%")
  ) %>%
  rename(Level = Age_Group, Count = n)

bmi_category_table <- fp5 %>%
  count(BMI_category) %>%
  mutate(
    Variable = "BMI_category",
    Proportion = paste0(round(100 * n / sum(n), 1), "%")
  ) %>%
  rename(Level = BMI_category, Count = n)

genhlth_binary_table <- fp5 %>%
  count(GENHLTH_binary) %>%
  mutate(
    Variable = "GENHLTH_binary",
    Proportion = paste0(round(100 * n / sum(n), 1), "%")
  ) %>%
  rename(Level = GENHLTH_binary, Count = n)

physical_activity_table <- fp5 %>%
  count(Physical_Activity) %>%
  mutate(
    Variable = "Physical_Activity",
    Proportion = paste0(round(100 * n / sum(n), 1), "%")
  ) %>%
  rename(Level = Physical_Activity, Count = n)

# Combine the individual tables using rbind()
table_combined <- rbind(
  age_group_table,
  bmi_category_table,
  genhlth_binary_table,
  physical_activity_table
)

# Reorder columns to Variable | Level | Count | Proportion
table_combined <- table_combined %>%
  select(Variable, Level, Count, Proportion)

# Print the final table to check outcome
print(table_combined, n = Inf)

```

```{r}
#Export output as csv file named Table1
write.csv(table_combined, "Table1.csv", row.names = FALSE)
```

```{r}
#Output the final sample selection process
flowchart <- grViz(diagram = "digraph flowchart{ # gives beginning of flowchart

      node [fontname = Calibri, shape = rectangle, fontsize=10] 
      
      node1 [label = '@@1']
      node2 [label = '@@2'] 
      node3 [label = '@@3'] 
      node4 [label = '@@4'] 
      node5 [label = '@@5'] 

      node1 -> node2 -> node3 -> node4 -> node5
}
      [1]: 'Initial records from BRFSS n = 433,323'
      [2]: 'Excluding 7,779 individuals with age entries: \\n Do not know/Refused/Missing n = 425,544'
      [3]: 'Excluding 47,527 individuals in age categories \\n that are < 30 n = 378,017'
      [4]: 'Excluding 126,402 individuals with no entries \\n in BMI Category and Physical Activity n = 251,615'
      [5]: 'Excluding 446 individuals with entries in their \\n general health status that include Refused, Not sure, Do not know n = 251,169'
      ")

flowchart
```

```{r}
#Export the flkowchart as a png file named flowchart_large with the dimensions as shown
export_svg(flowchart) %>% 
  charToRaw() %>% 
  rsvg_png("flowchart_large.png", width = 1600, height = 1400)
```


```{r}
<<<<<<< HEAD
=======
library(broom)
>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
#Simple logistic regression model to assess the association between physical activity and self-reported excellent health status.

model_a <- glm(GENHLTH_binary ~ Physical_Activity, 
               data = fp5, 
               family = binomial)

summary(model_a)

odds_ratio <- exp(coef(model_a))
confint_odds <- exp(confint(model_a))  

results <- data.frame(
  Term = names(coef(model_a)),
  Estimate = coef(model_a),
  Odds_Ratio = odds_ratio,
  CI_Lower = confint_odds[, 1],
  CI_Upper = confint_odds[, 2],
  P_Value = coef(summary(model_a))[, "Pr(>|z|)"]
)

results

```
Individuals who engaged in physical activity "times per week" had 1.22 times higher odds of reporting excellent health compared to those who engaged in physical activity "times per month."

```{r}
# Including BMI category and age group as covariates to adjust for their potential confounding effects.
model_b <- glm(GENHLTH_binary ~ Physical_Activity + BMI_category + Age_Group,
               data = fp5,
               family = binomial)

model_summary <- summary(model_b)

output_table <- data.frame(
  Term = rownames(model_summary$coefficients),
  Estimate = model_summary$coefficients[, "Estimate"],
  Odds = exp(model_summary$coefficients[, "Estimate"]), 
  PercentageChange = (exp(model_summary$coefficients[, "Estimate"]) - 1) * 100, 
  StdError = model_summary$coefficients[, "Std. Error"],
  ZValue = model_summary$coefficients[, "z value"],
  PValue = model_summary$coefficients[, "Pr(>|z|)"]

)

# Saving the table to a CSV file
write.csv(output_table, "model_b_output.csv", row.names = FALSE)

output_table
```

```{r}
#Logistic regression with interaction to check for effect modification with BMI and age
model_bmi <- glm(GENHLTH_binary ~ Physical_Activity * BMI_category, 
                 data = fp5, family = binomial)

model_age <- glm(GENHLTH_binary ~ Physical_Activity * Age_Group, 
                 data = fp5, family = binomial)


```

```{r}
#Logistic regression model to assess the combined effects of physical activity, BMI, and age group

model_combined <- glm(GENHLTH_binary ~ Physical_Activity * BMI_category + Physical_Activity * Age_Group,
                      data = fp5,
                      family = binomial)

model_summary <- summary(model_combined)

conf_intervals <- confint(model_combined)

output_table <- data.frame(
  Term = rownames(model_summary$coefficients),
  Estimate = model_summary$coefficients[, "Estimate"],
  Odds = exp(model_summary$coefficients[, "Estimate"]),  
  CI_Lower = exp(conf_intervals[, 1]),  
  CI_Upper = exp(conf_intervals[, 2]),  
  PercentageChange = (exp(model_summary$coefficients[, "Estimate"]) - 1) * 100,  
  StdError = model_summary$coefficients[, "Std. Error"],
  ZValue = model_summary$coefficients[, "z value"],
  PValue = model_summary$coefficients[, "Pr(>|z|)"]
)

print(output_table)

write.csv(output_table, "model_combined_output_with_CI.csv", row.names = FALSE)


```

```{r}
#Compaing the AICs to determine the best model fit
anova(model_bmi, model_combined, test = "LRT")
AIC(model_bmi, model_combined)
```

```{r}
<<<<<<< HEAD
=======
if (!require(car)) install.packages("car")
library(car)

>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
# Checking chosen model for multicollinearity with VIF
vif_values <- vif(model_combined)

vif_values
```
Interpretation: There is no multicollinearity

```{r}
<<<<<<< HEAD
=======

if (!require("lmtest")) install.packages("lmtest")
library(lmtest)

>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
# Performing the Breusch-Pagan test on the model for heteroscadastiity
bp_test <- bptest(model_b)
bp_test

```
Interpretation: There was statistically significant heteroscedasticity in the model.

```{r}
#Checking for linearity assumption in the logistic regression model using a residuals vs fitted values plot

plot(model_combined$fitted.values, residuals(model_combined), 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

```
Interpretation: The choice model violates the linearity assumption.

```{r}
# Checking for model fit using the Hosmer-Lemeshow goodness-of-fit test

<<<<<<< HEAD
=======
install.packages("ResourceSelection") #I had not previously installed this package

library(ResourceSelection)

>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
hoslem.test(model_combined$y, fitted(model_combined), g = 10) 

```
#The statistically significant HLtest shows the poor model fit.

```{r}
# Recode Physical Activity into 1 and 2
fp5$Physical_Activity_Cat <- ifelse(fp5$Physical_Activity == "Times per week", 1,
                                    ifelse(fp5$Physical_Activity == "Times per month", 2, NA))

# Apply a log transform to the categorical variable
fp5$log_Physical_Activity_Cat <- log(fp5$Physical_Activity_Cat + 1)

# Updating the model with the new transformed variable
model_combined_log <- glm(
  GENHLTH_binary ~ log_Physical_Activity_Cat + BMI_category + Age_Group +
  log_Physical_Activity_Cat * BMI_category + log_Physical_Activity_Cat * Age_Group,
  data = fp5,
  family = binomial
)

summary(model_combined_log)

```

```{r}
#Rechecking for heteroscedasticity in the log-transformed model

bp_test <- bptest(model_combined_log) 
bp_test
```
Interpretation: The heteroscedasticity persists despite the log transformation. Hence, the heteroscedasticity likely arose from factors unrelated to the scale of the variables.

```{r}
# Predict probabilities
fp5$predicted_prob <- predict(model_combined, type = "response")

```

```{r}
<<<<<<< HEAD
# Faceted scatterplot to visualize output
ggplot(fp5, aes(x = Physical_Activity, y = predicted_prob, color = BMI_category)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ Age_Group, nrow = 3, scales = "free_y") +  
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
=======
library(ggplot2)

# Cleaned plot with better facet layout and spacing
ggplot(fp5, aes(x = Physical_Activity, y = predicted_prob, color = BMI_category)) +
  geom_point(alpha = 0.6) +  # Adjust transparency for clarity
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +  # Thinner smooth lines
  facet_wrap(~ Age_Group, nrow = 3, scales = "free_y") +  # Fewer rows for better spacing
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +  # Wrap long x-axis labels
>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
  labs(
    title = "Combined Interaction: Physical Activity, BMI, and Age Group",
    x = "Physical Activity",
    y = "Predicted Probability of Excellent Health",
    color = "BMI Category"
  ) +
  theme_minimal() +
  theme(
<<<<<<< HEAD
    strip.text = element_text(size = 10, face = "bold"),  
    axis.text.x = element_text(angle = 45, hjust = 1),  
    legend.position = "bottom"                           
=======
    strip.text = element_text(size = 10, face = "bold"),  # Larger facet labels
    axis.text.x = element_text(angle = 45, hjust = 1),   # Rotate x-axis text for clarity
    legend.position = "bottom"                           # Move legend to the bottom
>>>>>>> e6cb2824e110adc38d2396479dd2c8de20312dd0
  )
```

```{r}
#Exporting the faceted scatterplot as a png file
ggsave("interaction_plot.jpeg", plot = last_plot(), width = 10, height = 8, dpi = 300)

```